Rationality, Evolution \& the Environment

### 2015-08-20 paper/paper.tex
choice evol

\section{Results: the basic model}

@@

\subsection{Numerical simulations}

@@

Similarly to the evolutionary competition that we simulated before, we can study the evolutionary fitness of this extended set of players when they are all represented in the population. Table \ref{tab:Utils-MetaGame} shows the averge fitness of the enlarged competition.

\begin{table*}[ht]
\centering
\begin{tabular}{ccccccccc}
  \hline
 & $\tau^{reg}, \Delta(A_{-i})$ 
 & $\tau^{\pi}, \Delta(A_{-i})$ 
 & $\tau^{com}, \Delta(A_{-i})$
 & $\tau^{alt}, \Delta(A_{-i})$ \\ 
  \hline
  $\tau^{reg}, \Delta(A_{-i})$ & 6.663 & 6.662 & 5.829 & 7.105 \\ 
  $\tau^{\pi}, \Delta(A_{-i})$ & 6.486 & 6.484 & 6.088 & 6.703 \\ 
  $\tau^{com}, \Delta(A_{-i})$ & 6.323 & 6.758 & 5.496 & 6.977 \\ 
  $\tau^{alt}, \Delta(A_{-i})$ & 5.949 & 5.722 & 5.326 & 6.396 \\ 
   \hline
\end{tabular}
\caption{Average payoff in simulations of 100,000 symmetric $2 \times 2$ games}
\label{tab:Utils-MetaGame}
\end{table*}

The following proposition confirms this result.

@@

Finally, Table \ref{tab:Utils-MetaGame2} shows the average fitness when all the epistemic and
player types introduced so far are simultaneously present in the population.

Again, this shows that there are plausible and simple conditions under which
agents who represent the game in terms of regret may be favored by
evolutionary selection and that, more specifically, regret minimizers
are evolutionarily supported.

\begin{table*}[ht]
\centering
\begin{tabular}{ccccccccc}
  \hline
 & $\tuple{\tau^{reg}, \Delta(A)}$ 
 & $\tuple{\tau^{\pi}, \Delta(A)}$ 
 & $\tuple{\tau^{com}, \Delta(A)}$
 & $\tuple{\tau^{alt}, \Delta(A)}$
 & $\tuple{\tau^{reg}, \bar{\mu}}$ 
 & $\tuple{\tau^{\pi}, \bar{\mu}}$ 
 & $\tuple{\tau^{com}, \bar{\mu}}$
 & $\tuple{\tau^{alt}, \bar{\mu}}$ \\ 
  \hline
  $\tuple{\tau^{reg}, \Delta(A)}$ &  6.663 & 6.662 & 5.829 & 7.105 & 6.663 & 6.663 & 5.829 & 7.489 \\  
  $\tuple{\tau^{\pi}, \Delta(A)}$ &  6.486 & 6.484 & 6.088 & 6.703 & 6.486 & 6.486 & 6.088 & 6.875 \\  
  $\tuple{\tau^{com}, \Delta(A)}$ &  6.323 & 6.758 & 5.496 & 6.977 & 6.323 & 6.323 & 5.496 & 7.149 \\  
  $\tuple{\tau^{alt}, \Delta(A)}$ &  5.949 & 5.722 & 5.326 & 6.396 & 5.949 & 5.949 & 5.326 & 6.568 \\  
  $\tuple{\tau^{reg}, \bar{\mu}}$ &  6.663 & 6.662 & 5.829 & 7.105 & 6.663 & 6.663 & 5.829 & 7.489 \\  
  $\tuple{\tau^{\pi}, \bar{\mu}}$ &  6.663 & 6.662 & 5.829 & 7.105 & 6.663 & 6.663 & 5.829 & 7.489 \\  
  $\tuple{\tau^{com}, \bar{\mu}}$ &  6.323 & 6.758 & 5.496 & 6.977 & 6.323 & 6.323 & 5.496 & 7.149 \\  
  $\tuple{\tau^{alt}, \bar{\mu}}$ &  6.331 & 5.893 & 5.497 & 6.566 & 6.331 & 6.331 & 5.497 & 7.152 \\   
   \hline                          
\end{tabular}                      
\caption{Average payoff in simulations of 100,000
  randomly generated $2 \times 2$ symmetric games.}
\label{tab:Utils-MetaGame2}        
\end{table*}

@@

>>>>

\section{Results: the basic model}
\label{sec:results:-basic-model}

@@

\subsection{Numerical simulations}
\label{sec:numer-simul}

@@

Similarly to the evolutionary competition that we simulated before, we can study the
evolutionary fitness of this extended set of players when they are all represented in the
population. Table~\ref{tab:ExpectedFitness_2x2_Full} shows simulation results that approximate the
expected fitness in the relevant meta-game.

\begin{table}[]
\centering
\footnotesize
\begin{tabular}{ccccccccc}
  \hline
 & $\tuple{\tau^{reg}, \Delta(A)}$ 
 & $\tuple{\tau^{\pi}, \Delta(A)}$ 
 & $\tuple{\tau^{com}, \Delta(A)}$
 & $\tuple{\tau^{alt}, \Delta(A)}$
 & $\tuple{\tau^{reg}, \bar{\mu}}$ 
 & $\tuple{\tau^{\pi}, \bar{\mu}}$ 
 & $\tuple{\tau^{com}, \bar{\mu}}$
 & $\tuple{\tau^{alt}, \bar{\mu}}$ \\ 
  \hline
  $\tuple{\tau^{reg}, \Delta(A)}$ & 6.663 & 6.662 & 5.829 & 7.105 & 6.663 & 6.663 & 5.829 & 7.489 \\
  $\tuple{\tau^{\pi}, \Delta(A)}$ & 6.486 & 6.484 & 6.088 & 6.703 & 6.486 & 6.486 & 6.088 & 6.875 \\
  $\tuple{\tau^{com}, \Delta(A)}$ & 6.323 & 6.758 & 5.496 & 6.977 & 6.323 & 6.323 & 5.496 & 7.149 \\
  $\tuple{\tau^{alt}, \Delta(A)}$ & 5.949 & 5.722 & 5.326 & 6.396 & 5.949 & 5.949 & 5.326 & 6.568 \\
  $\tuple{\tau^{reg}, \bar{\mu}}$ & 6.663 & 6.662 & 5.829 & 7.105 & 6.663 & 6.663 & 5.829 & 7.489 \\
  $\tuple{\tau^{\pi}, \bar{\mu}}$ & 6.663 & 6.662 & 5.829 & 7.105 & 6.663 & 6.663 & 5.829 & 7.489 \\
  $\tuple{\tau^{com}, \bar{\mu}}$ & 6.323 & 6.758 & 5.496 & 6.977 & 6.323 & 6.323 & 5.496 & 7.149 \\
  $\tuple{\tau^{alt}, \bar{\mu}}$ & 6.331 & 5.893 & 5.497 & 6.566 & 6.331 & 6.331 & 5.497 & 7.152 \\
   \hline                          
\end{tabular}                      
\caption{Average payoff in simulations of 100,000
  randomly generated $2 \times 2$ symmetric games.}
\label{tab:ExpectedFitness_2x2_Full}        
\end{table}

\todo[inline]{is it true that competitive preferences types are behaviorally equivalent under
  either epistemic type? can we show that?}

The approximations in Table~\ref{tab:ExpectedFitness_2x2_Full} confirm basic
intuitions. Averaging over randomly sampled games, everybody would like to have an altruistic
opponent and nobody would like to play against a competitive opponent. Moreover, altruistic
types come up strictly dominated by competitive types, but competitive types themselves are
always worse than any of the behaviorally equivalent types $(\tau^{reg}, \Delta(A_{-i}))$,
$(\tau^{\pi}, \overline{\mu})$, and $(\tau^{reg}, \overline{\mu})$, except against maximin
players $\tuple{\tau^{\pi}, \Delta(A)}$. This is a noteworthy results in the light of the fact
that evolving altruistic preferences have been shown to support cooperative behavior in a
single stage game \myalert{[CITE]}.\todo{@Paolo: check whether this is true, please!} In
contrast, averaging over payoffs in multiple stage games, like we do here, makes altruistic
preferences prime victims of evolutionary eradication.

It is easy to see then that the previous result still obtains for the larger meta-game in
Table~\ref{tab:ExpectedFitness_2x2_Full}: $(\tau^{reg}, \Delta(A_{-i}))$,
$(\tau^{\pi}, \overline{\mu})$, and $(\tau^{reg}, \overline{\mu})$ are still neutrally stable;
all simulations of the (discrete-time) replicator dynamic end in populations consisting of only
these types in arbitrary proportion.

The following proposition confirms this result.\todo{rephrase to acknowledge all epistemic types}

@@

\begin{proof}
See Appendix.\todo{is this there already?}
\end{proof}

In sum, the presence of other plausible preference types, such as competitive and altruistic
types does not undermine the previous results about the evolutionary drive towards regret-based
subjective preferences.

@@

Results from Section~\ref{sec:results:-basic-model} relied heavily on
Fact~\ref{fact:equivalence2x2} that regret minimization $\tuple{\tau^{reg}, \Delta(A)}$ is
behaviorally equivalent with expected regret minimization $\tuple{\tau^{reg}, \bar{\mu}}$ and
expected utility maximization $\tuple{\tau^{\pi}, \bar{\mu}}$. This is no longer true when we
look at arbitrary $n \times n$ games with $n \ge 2$.

Table~\ref{tab:ExpectedFitness_10x10} therefore gives approximations for expected fitness in a
meta-game over a class over a class of $n \times n$ symmetric games where $n$ is randomly drawn
from $\set{2, \dots, 10}$. Concretely, the numbers in Table~\ref{tab:ExpectedFitness_10x10} are
averages of payoffs obtained in 100,000 randomly sampled games, where each game was sampled by
first picking a number of acts $n \in \set{2, \dots, 10}$ uniformly at random, and then filling
the resulting $n \times n$ payoff matrix of the stage game with i.i.d.~sampled payoffs as
before.

\begin{table}[]
\centering
\footnotesize
\begin{tabular}{ccccccccc}
  \hline
 & $\tuple{\tau^{reg}, \Delta(A)}$ 
 & $\tuple{\tau^{\pi}, \Delta(A)}$ 
 & $\tuple{\tau^{com}, \Delta(A)}$
 & $\tuple{\tau^{alt}, \Delta(A)}$
 & $\tuple{\tau^{reg}, \bar{\mu}}$ 
 & $\tuple{\tau^{\pi}, \bar{\mu}}$ 
 & $\tuple{\tau^{com}, \bar{\mu}}$
 & $\tuple{\tau^{alt}, \bar{\mu}}$ \\ 
  \hline
  $\tuple{\tau^{reg}, \Delta(A)}$ & 6.567 & 6.570 & 5.650 & 6.992 & 6.564 & 6.564 & 5.593 & 7.409 \\
  $\tuple{\tau^{\pi}, \Delta(A)}$ & 6.476 & 6.483 & 5.896 & 6.818 & 6.484 & 6.484 & 5.850 & 7.124 \\
  $\tuple{\tau^{com}, \Delta(A)}$ & 6.468 & 6.647 & 5.512 & 7.169 & 6.578 & 6.578 & 5.577 & 7.354 \\
  $\tuple{\tau^{alt}, \Delta(A)}$ & 5.968 & 5.923 & 5.363 & 6.685 & 5.975 & 5.975 & 5.086 & 6.973 \\
  $\tuple{\tau^{reg}, \bar{\mu}}$ & 6.908 & 6.918 & 5.988 & 7.456 & 6.929 & 6.929 & 5.934 & 7.783 \\
  $\tuple{\tau^{\pi}, \bar{\mu}}$ & 6.908 & 6.918 & 5.988 & 7.456 & 6.929 & 6.929 & 5.934 & 7.783 \\
  $\tuple{\tau^{com}, \bar{\mu}}$ & 6.529 & 6.680 & 5.445 & 7.276 & 6.542 & 6.542 & 5.521 & 7.440 \\
  $\tuple{\tau^{alt}, \bar{\mu}}$ & 6.450 & 6.337 & 5.772 & 6.978 & 6.457 & 6.457 & 5.479 & 7.500 \\
   \hline                          
\end{tabular}                      
\caption{Average payoff in simulations of 100,000
  randomly generated $n \times n$ symmetric games with $n$ randomly drawn from $\set{2, \dots, 10}$.}
\label{tab:ExpectedFitness_10x10}        
\end{table}
